Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-09-22 manu_draft.Rmd
adding some intro text building off of PDV text

>>>>

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.  The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.  At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that are not of direct interest for longer-term trends.  For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends. However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.  Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.

In this paper we propose analysis methods that separate the signal-extraction of fine-scale data from the trend analysis at longer time-scales, with full propagation of statistical uncertainties between the two analysis stages.  For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines.  While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data, alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines. Analysis questions or management goals that require information about about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.  For the trend-analysis stage, error propagation in summary statistics can be addressed using mixed-effects meta-analysis, enabling the application of established tools that incorporate uncertainties in model estimates of summary statistics.  This two-stage approach is useful because across-year time-scales of interest can be imposed or estimated explicitly, which contrasts with implicit absorption of time-scales into spline formulations in other methods.  For example, across-year smoothing methods can be applied to seasonal averages from the signal-extraction stage while accurately propagating uncertainties.  On a practical level, the signal-extraction stage can be viewed as creating a "data product" that can be used in second-stage analyses without the difficulty of embedding every long-term trend question in a GAM modeling framework to model the fine-scale fluctuations.

The smoothing methods under discussion here are distinct from older methods of trend detection, such as seasonal Kendall tests or other non-parametric approaches.  The former has been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years.  The limitations are such that only the magnitude and direction of a change between selected a priori years can be identified, nor are any quantitative descriptions of changes occurring at different temporal scales explicitly provided.  Alternative approaches that decompose a time series into additive or multiplicative components have also been used, such as the seasonal time series composition that isolates a trend, seasonal, and residual component from an observed time series.  Although computationally efficient, such approaches may over-simplify trends into absolute components that do not smoothly vary over time, nor do they allow for explicit hypothesis testing.  More recently, moving window regression approaches have been applied to dynamically estimate parameters that influence temporal changes in water quality time series.  The Weighted Regression on Time, Discharge, and Season (WRTDS) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.  The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable.

GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.  The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.  Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity.  Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic or management questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).  Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.

Although our approach is not the first to apply GAMs to environmental monitoring data, our approach is distinct from previous applications in several ways.  First, some approaches attempt to estimate a parameter such as a seasonal average with the GAM itself [q1].  This represents a mismatch between time scales and does not make sense as an estimation method if there is really no seasonal treatment factor, for example.  Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the GAM fitting machinery [q2].  There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be hard to interpret.  Perhaps more important, it risks obscuring the different time-scales implicit in choices such as number of spline knots, leaving the rendering results harder to interpret.  Third…[q3]...  We illustrate each of these points in more detail below.

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Measurements from water samples include chlorophyll a, [others].  Analysis is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data could be misleading.  Examples of long-term trend questions include: Are there significant trends in spring mean chl-a at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chl-a levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chl-a or in off-bloom baseline levels?  We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.  The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evlauating these trends between year.  Practically, this approach answers if a seasonal change has occurred across years of interest by providing an estimate of the change and its associated confidence as propagated through the GAM.
